Decade of Lost Precision: A General Framework For Image Feature Matching Without Geometric Constraints

### 2014-04-21 text/journal/paper.tex
Last few changes before submitting

\begin{abstract}
%\boldmath
Computer vision applications that involve the matching of local image features frequently use \emph{Ratio-Match} as introduced by Lowe and others, but is this really the optimal approach?  We formalize the theoretical foundation of \emph{Ratio-Match} and propose a general framework encompassing \emph{Ratio-Match} and three other matching methods. Using this framework, we establish a theoretical performance ranking in terms of precision and recall, proving that all three methods consistently outperform or equal \emph{Ratio-Match} in cases achieving a 20 percentage-point performance increase.

We confirm the theoretical results experimentally on over 3000 image pairs and show that we can increase matching precision without further assumptions about the images we are using.  These gains are achieved by making only a few key changes of the \emph{Ratio-Match} algorithm and do not increase computation times.
\end{abstract}

@@

\section{Introduction}
%
Matching image points is a crucial ingredient in almost all computer vision algorithms that deal with sparse local image features. This includes image categorization \cite{bosch2008scene}, image stitching \cite{brown2007automatic}, object detection \cite{zhang2007local}, and near duplicate detection \cite{zhao2009scale}, to mention just a few examples.  These problems all rely on being able to accurately find the correspondence(s) of a point on an object in a \emph{query image} given one or more \emph{target images} that might contain the same object.  In many applications the \emph{target images} have undergone transformations with respect to the \emph{query image}; in stereo vision, the viewpoint is different, while in object recognition and near duplicate  detection both the lighting and even the object itself may also be transformed.

In the literature two approaches to feature point matching have been pursued and later merged. The geometric approach tries to find unique keypoints in an image pair and match them based on their location.  The descriptor centric approach on the other hand uses the local image information around the keypoint to create a descriptor and matches descriptors based on their similarity.

@@

Brown and Lowe \cite{brown2005multi} extend ratio match to deal with a set of images by using not the ratio of the best and second best correspondence, but the average ratio of the best and the average of second best correspondences across a set of images.  Rabin et al.\ tries to enhance descriptor matching by looking at the statistical distribution of local features in the matched images, and only return a match when such a correspondence would not occur by mere chance \cite{rabin2009statistical}. Finally, a precursor of the algorithms discussed in this paper was introduced by the authors as \emph{Mirror-Match} \cite{arnfred2013mirror}, which makes use of the feature points in both images to decide if a match is valid.

A plethora of solutions have combined ratio match with various geometric constraints to improve matching. These constraints are based on assumptions regarding the transformation between the query and \emph{target images}. At the stricter end we have epipolar constraints assuming that the images can be tied by a homography \cite{torr2000mlesac, chum2005matching} and angular constraints assuming the correspondences are angled similarly \cite{kim2008efficient, schmid1997local}. Often these approaches are made computationally feasible by modelling the case of feature correspondences as an instance of graph matching where each feature is a vertex and edge values correspond to a geometric relation between two features. In this case approximate graph matching algorithms can be used to efficiently establish an isomorphism between the graph of features in two images \cite{leordeanu2005spectral, torresani2008feature, yarkony2010covering}. Finally others define image regions and reject or accept correspondences based on the regions they connect \cite{cho2009feature, wu2011robust}.

Any matching method relying on geometric constraints is restrained by inherent assumptions about the geometric relationship between the two images. Broad assumptions such as the epipolar constraint only apply in simple image transformations. For more complex transformations we need models suitable for each particular case, which limits the algorithm to the subset of images that fit the model.  In the case of object recognition for example, the transformations from one scene to another often feature a change in perspective, background, and sometimes variations within the object itself: a person can change pose, a car model can have different configurations, a flower can bloom etc.

@@

In the author's last paper we introduced \emph{Mirror-Match} and \emph{Mirror-Match with Clustering}, two algorithms that we showed to perform better than the state of the art. The novel contributions of this paper consists of two new algorithms completing a framework of four related matching algorithms as well as a theoretical foundation for comparing the algorithms. We use this foundation to formally prove a ranking in terms of performance for the different algorithm which enables us to understand why \emph{Mirror-Match} performs better than \emph{Ratio Match} in the first place. In addition we test these results on a much larger dataset consisting of over 3000 image pairs.

The paper is organized as follows. I section~\ref{S:MatchingMethods} we introduce the original \emph{Ratio-Match}, extend on this method to introduce the proposed framework and go on to compare the members of the framework theoretically.  In section~\ref{S:Experiments} we present evaluations and discuss the results obtained.  Section~\ref{S:Summary} concludes the paper.

@@

To make further discussion easier we introduce the following nomenclature:
\begin{itemize}
\item{Let $f_q$ be any feature point in the \emph{query image} that we 
    are currently trying to match}.
\item{Let $F$ be a set of features. $F_{t}$ for example is the set 
    of features from the \emph{target image}}.
\item{Let $\tau \in [0 \ldots 1]$ be a threshold used to decide whether 
    to keep a match or discard it. We keep a match only if is has a 
\emph{uniqueness ratio} lower than $\tau$}
\item{Let the \emph{proposed match} be the nearest neighbor of the query 
    feature $f_q$ picked from a set of feature points that we will call 
the \emph{proposal set} which does not contain the query feature itself}
\item{Let the \emph{baseline match} be the nearest neighbor of the query 
    feature $f_q$ picked from a set of feature points that we will call 
the \emph{baseline set} which neither contains the \emph{proposed match} 
nor the query feature}
\end{itemize}

@@

Given two images we can construct the proposal and baseline sets in six different ways as illustrated in Figure~\ref{fig:overview}. In the case of \emph{Ratio-Match} \cite{lowe2004sift} the \emph{proposal set} consists of all the feature points in the \emph{target image}.  Similarly the \emph{baseline set} consists of all the feature points in the \emph{target image} except the \emph{proposed} match. For \emph{Ratio-Match} this practically means that given a query feature, $f_q$, the \emph{proposed match} and the \emph{baseline match} are the two nearest neighbors in the set of feature points in the \emph{target image} and the ratio is decided uniquely based on the \emph{target image}. In turn \emph{Mirror-Match} as introduced by the authors in \cite{arnfred2013mirror}, works by constructing the \emph{proposal set} and \emph{baseline set} by combining features from both the \emph{query} and \emph{target image}. By similar variations of the \emph{proposal} and \emph{baseline set} we derive four more algorithms as illustrated in Figure~\ref{fig:overview}, however we will not discuss \emph{Self-Match-Ext} and \emph{Both-Match} further in this paper since we show them to be equivalent with \emph{Self-Match} and \emph{Mirror-Match} respectively in \ref{A:self} and \ref{A:mirror}.

@@

For some methods it is possible that the \emph{proposal match} is a feature from the \emph{query image}, in which case we discard the match as an untrue correspondence.  It also happens that we encounter correspondences with a \emph{uniqueness ratio} above $1$, in which case the match is also discarded.  Take for example the case of \emph{Self-Match} where we might find that the nearest neighbor of a feature in the \emph{target image} is further from the query feature than the nearest neighbor in the \emph{query image}. In this case the match is discarded.

@@

The main distinguishing factor in between the algorithms in the framework is the final ratio between the distance of the two nearest neighbors of a query feature which decides if we keep or discard a match. We illustrate this process in Figure~\ref{fig:matching}.

We can define a calculation of this ratio common to all algorithms in the framework which allows us to compare the algorithms theoretically. To do so we first define the concept of a nearest neighbor and then based on that the \emph{uniqueness ratio}, $r$:

Given a feature, $f_i$ and a set of features, $F$ the nearest neighbor of $f_i$ in $F$ is calculated as follows: 
\begin{equation*}
\argmin_{f_j \in F} d(f_i, f_j)
\end{equation*}
Here, $d(f_i, f_j)$ is the distance between feature descriptors.With SIFT and SURF this is the Euclidean distance of the feature vectors \cite{lowe2004sift,bay2006surf}, whereas BRIEF, BRISK, and FREAK use the Hamming distance \cite{leutenegger2011brisk,calonder2010brief,alahi2012freak}.

@@

\begin{enumerate}
    \item{For a query and \emph{target image} we assume that for any point in 
        the \emph{query image} there is at most one real correspondence in the 
    \emph{target image} and no real correspondences in the \emph{query image} itself 
as assumed by \emph{Ratio-Match}.}
    \item{We assume that distances of feature descriptors are well 
            behaved.  More precisely, given $f_q$, a feature from the 
            \emph{query image}, and $f_{match}$, the real correspondence 
            from the set of features in the \emph{target image} 
            $F_{t}$, we have $\forall f_i \in F_{proposal}, 
            f_i \neq f_{match}: d(f_q,f_{match}) < d(f_q, f_i)$.}
    %\item{We assume that features within one image resemble each other 
            %more than features in different images. That is, given 
            %$f_a,
    %        f_i \in F_1$ and $f_b, f_j \in F_2$ where $(f_i, f_j)$ and 
    %    $(f_a, f_b)$ are true correspondences, and $F_1$ and $F_2$ are 
    %features collected from different images then we expect that 
            %$d(f_a, f_i)% < d(f_a, f_j)$ and $d(f_b, f_j) < d(f_b, 
            %f_i)$}
     \item{For a set of query features $F_{q}$ with a true 
             correspondence in the \emph{target image}, we assume that 
             the distribution of \emph{uniqueness ratios} using $F_{t}$ as 
             the baseline set is similar to using $F_{q}$, since the 
         two images in the case of true correspondences are guaranteed 
     to share part of the same scene.}
    \end{enumerate}

@@

Next we show that \emph{Mirror-Match} is equal or better than \emph{Ratio-Match-Ext} in terms of precision and recall. Consider as before the nearest neighbor $f_{p}$ of a query feature $f_{q}$. When $f_{p}$ resides in the \emph{query image} both algorithms behave alike and discard the match.  However, consider the uniqueness ratio of \emph{Mirror-Match} for the case where $f_{p}$ resides in the \emph{target image}:
\begin{align*}
    r &= \text{r}(f_{q}, F_{p}, F_{b}) \\
        &= \text{r}(f_{q}, F_{p}, F_{q} \cup F_{t})\\
        &= max( \text{r}(f_{q}, F_{p}, F_{t}), 
    \text{r}(f_{q}, F_{p}, F_{q}) ).
\end{align*}
For a true correspondence the \emph{uniqueness ratio} using $F_{q}$ as a baseline is distributed similarly to the ratio when using $F_{t}$ according to assumption \#3. In this case the algorithm performs like \emph{Ratio-Match-Ext}.  However, for an untrue correspondence with a baseline match in the \emph{query image} which is closer than the baseline match in the \emph{target image}, \emph{Mirror Match} will return a worse \emph{uniqueness ratio}. This in turns means that \emph{Mirror Match} is equal or better than \emph{Ratio-Match-Ext} in terms of  \emph{Precision} while maintaining equal \emph{Recall}.

Using a similar procedure we prove that \emph{Self-Match} is equal to \emph{Self-Match-Ext} and \emph{Both-Match} is equal to \emph{Mirror Match} in \ref{A:self} and \ref{A:mirror} respectively.

@@

The first assumption states that there is at most one unique match to every feature point, which is often the case for natural images. However, for the recognition of object classes or in generated images with repetitive content, a point in the \emph{query image} might have several true correspondences in the \emph{target image}, so using a \emph{baseline match} from the \emph{target image} will lead to a much higher \emph{uniqueness ratio}. However, as long as the \emph{query image} does not contain repetitive objects, we can still use \emph{Self-Match} without loss of precision.

@@

For the experiments we use the \emph{SIFT} descriptor and keypoint detector with default parameters as implemented in the OpenCV library 2.4.6.  To further test how the framework performs across descriptors, we evaluate SURF \cite{bay2006surf}, BRISK \cite{leutenegger2011brisk}, BRIEF \cite{calonder2010brief}, and FREAK \cite{alahi2012freak} in Section~\ref{label:desc}.

@@

\subsection{Experiments}
\label{S:3dobjects}

@@

\begin{figure}[htb]
    \centering
    \begin{subfigure}[t]{0.22\columnwidth}
        \centering
        \includegraphics[width=\columnwidth]{images/Frog_A}
        \caption{Query Image}
        \label{fig:frog_a}
    \end{subfigure}%
    \quad %add desired spacing between images, e. g. ~, \quad, \qquad a 
    %blank line to force the subfigure onto a new line)
    \begin{subfigure}[t]{0.22\columnwidth}
        \centering
        \includegraphics[width=\columnwidth]{images/Frog_B}
        \caption{Auxiliary Image}
        \label{fig:frog_b}
    \end{subfigure}%
    \quad %add desired spacing between images, e. g. ~, \quad, \qquad 
    %a blank line to force the subfigure onto a new line)
    \begin{subfigure}[t]{0.22\columnwidth}
        \centering
        \includegraphics[width=\columnwidth]{images/Frog_C}
        \caption{Target Image}
        \label{fig:frog_c}
    \end{subfigure}%
    \quad %add desired spacing between images, e. g. ~, \quad, \qquad 
    %blank line to force the subfigure onto a new line)
    \begin{subfigure}[t]{0.22\columnwidth}
        \centering
        \includegraphics[width=\columnwidth]{images/Frog_C_lines}
        \caption{Target Image with epipolar constraints}
        \label{fig:frog_c_lines}
    \end{subfigure}%
    \caption{Creating epipolar constraints based on three source images \cite{moreels2007evaluation}.  (a) \emph{Query image} marked with the position of the feature we are attempting to match. (b) Auxiliary image, taken at the same rotation as the \emph{query image} but from a higher viewpoint. The line going through the image is the epipolar line of the feature point in the \emph{query image}.  The markers indicate all feature points in the image found near the epipolar line.  In c) we show the \emph{target image} which in this case is rotated 45 degrees from the \emph{query image}.  The line overlaid on the image represents the epipolar line corresponding to the feature point shown in the \emph{query image}.  The markers indicate all feature points in the image found near the epipolar line. In d) we show the \emph{target image} again, this time overlaid with the epipolar lines corresponding to all the features shown in the \emph{auxiliary image}. A true correspondence should be found within a small distance of one of the intersections of the line in c) and the lines in d). In this particular case both feature points shown in c) and d) could possible be a correct correspondence.
}%
    \label{fig:frog}%
\end{figure}%

@@

We evaluate all matching algorithms from our framework on the 3D objects dataset by matching images at different angular intervals. For each object we pick the \emph{query} image as the image taken at 10 degrees rotation for calibration stability.  We then match this image with the same object turned an additional $\Delta$ degrees, $\Delta \in \{5, 10, \ldots, 60\}$.  For every angle interval we compare images taken under 3 different lighting conditions as provided by the dataset. We include all objects in the database for which photos at 5 degree angle intervals are available except for the ``Rooster'' and ``Sponge'' objects due to image irregularities. For each plot we show the accumulated results on all 3D objects, weighted by the number of possible true correspondences for the particular object. This ensures that each object contributes equally to the final result, despite some objects resulting in disproportionally more matches than others.

@@

Figure~\ref{fig:all_objects} shows the performance of the different matching methods in our proposed framework for $12$ increasingly bigger angle differences. The results are shown in a precision / recall plot to make it easy to compare performance in terms of precision at similar levels of recall.  \emph{Ratio-Match} and \emph{Self-Match} show similar results, while \emph{Mirror-Match} and \emph{Ratio-Match-Ext} outperform both of them, showing the advantage of composing the \emph{proposed set} of features from both images. \emph{Mirror-Match} fares slightly but consistently better than \emph{Ratio-Match-Ext}. In general we see the largest performance improvements at lower recall and a  convergence of precision at higher recall. This is expected since a higher recall is a direct consequence of a more lenient threshold; as we let the threshold approach $1$, we lose the benefits of thresholding on the \emph{uniqueness ratio}, and all methods start approximating the results of a simple nearest neighbor match. For all but \emph{Ratio-Match}, the features with better matches within the same image are still weeded out, which explains the tail end of \emph{Ratio-Match} at high recall where the other algorithms no longer have results. As we demonstrated in \cite{arnfred2013mirror}, the removal of within-image matches improves performance on image pairs with partial or no overlap. For no overlap we would expect a matching algorithm to reject matches within regions that have no matching counterparts in the other image.

@@

To summarize, the methods perform as predicted on the 3D Objects dataset with \emph{Mirror-Match} in general performing better than \emph{Ratio-Match-Ext}, which in turn outperforms \emph{Ratio-Match}. The experimental results also show that \emph{Self-Match} overall perform equal to or better than \emph{Ratio-Match}.

\subsection{Descriptor Evaluation}
\label{label:desc}

@@

The descriptors were evaluated on 15 images of the 3D object dataset with a fixed angle offset of 25 degrees (pictured in Figure~\ref{fig:3d_objects}).  Otherwise the evaluation performed in the same manner as the general evaluation of 3D Objects in Section~ref{S:3dobjects}, and the data presented is accumulated over the 15 objects, each set of results weighted by the possible number of correspondences.

The results in Figure~\ref{fig:descriptors} are again in line with the theoretical performance analysis of Section~\ref{S:Proofs} . They also show a clear difference in the variance of matching method performance between SIFT and the rest of the descriptors. The performance of \emph{Mirror-Match} using SIFT shows a clear increase in precision over \emph{Ratio-Match} at equal recall levels. For other descriptors the improvement is much less pronounced.  For SURF, BRISK, and BRIEF we see a small improvement in precision using \emph{Mirror-Match} over \emph{Ratio-Match} and \emph{Ratio-Match-ext}, whereas FREAK yields similar results no matter what algorithm is used.  For all descriptors \emph{Self-Match} performs on par with \emph{Ratio-Match}.

@@

\subsection{Performance}

@@

From an evaluation using images of rotated 3D objects, we have shown that the theoretical conclusions are reflected in the experimental results with \emph{Mirror-Match} often outperforming \emph{Ratio-Match} significantly over 3024 image pairs. Finally we show that the theoretical conclusions hold across a variety of popular feature descriptors (SIFT, SURF, BRISK, BRIEF, and FREAK). These performance gain come for free in terms of both computational complexity as well as actual computing time.

>>>>

\begin{abstract}
%\boldmath
Computer vision applications that involve the matching of local image features frequently use \emph{Ratio-Match} as introduced by Lowe and others, but is this really the optimal approach?  We formalize the theoretical foundation of \emph{Ratio-Match} and propose a general framework encompassing \emph{Ratio-Match} and three other matching methods. Using this framework, we establish a theoretical performance ranking in terms of precision and recall, proving that all three methods consistently outperform or equal \emph{Ratio-Match}.

We confirm the theoretical results experimentally on over 3000 image pairs and show that we can increase matching precision by up to 20 percentage-points without further assumptions about the images we are using.  These gains are achieved by making only a few key changes of the \emph{Ratio-Match} algorithm and do not increase computation times.
\end{abstract}

@@

\section{Introduction}
%
Matching image points is a crucial ingredient in almost all computer vision applications that deal with sparse local image features, such as image categorization \cite{bosch2008scene}, image stitching \cite{brown2007automatic}, object detection \cite{zhang2007local}, and near duplicate detection \cite{zhao2009scale}, to mention just a few examples.  All of these rely on accurately finding the correspondence(s) of a point on an object in a \emph{query image} given one or more \emph{target images} that might contain the same object.  In many applications the \emph{target images} have undergone transformations with respect to the \emph{query image}; in stereo vision, the viewpoint is different, while in object recognition and near duplicate  detection both the lighting and even the object itself may also be transformed.

In the literature two approaches to feature point matching have been pursued and later merged. The geometric approach tries to find unique keypoints in an image pair and matches them based on their location.  The descriptor centric approach on the other hand uses the local image information around the keypoint to create a descriptor and matches descriptors based on their similarity.

@@

Brown and Lowe \cite{brown2005multi} extend ratio match to deal with a set of images by using not the ratio of the best and second best correspondence, but the average ratio of the best and the average of second best correspondences across a set of images.  Rabin et al.\ \cite{rabin2009statistical} try to enhance descriptor matching by looking at the statistical distribution of local features in the matched images, and only return a match when such a correspondence would not occur by mere chance. Finally, a precursor of the algorithms discussed in this paper was introduced by the authors as \emph{Mirror-Match} \cite{arnfred2013mirror}, which makes use of the feature points in both images to decide if a match is valid.

A plethora of solutions have combined ratio match with various geometric constraints to improve matching. These constraints are based on assumptions regarding the transformation between the query and \emph{target images}. At the stricter end we have epipolar constraints assuming that the images can be tied by a homography \cite{torr2000mlesac, chum2005matching} and angular constraints assuming the correspondences are angled similarly \cite{kim2008efficient, schmid1997local}. Often these approaches are made computationally feasible by modeling the case of feature correspondences as an instance of graph matching, where each feature is a vertex, and edge values correspond to a geometric relation between two features. In this case approximate graph matching algorithms can be used to efficiently establish an isomorphism between the graph of features in two images \cite{leordeanu2005spectral, torresani2008feature, yarkony2010covering}. Others define image regions and reject or accept correspondences based on the regions they connect \cite{cho2009feature, wu2011robust}.

Any matching method relying on geometric constraints is limited by inherent assumptions about the geometric relationship between the two images. Broad assumptions such as the epipolar constraint only apply in simple image transformations. For more complex transformations we need models suitable for each particular case, which limits the algorithm to the subset of images that fit the model.  In the case of object recognition for example, the transformations from one scene to another often feature a change in perspective, background, and sometimes variations within the object itself: a person can change pose, a car model can have different configurations, a flower can bloom etc.

@@

In our previous paper \cite{arnfred2013mirror}, we introduced \emph{Mirror-Match} and \emph{Mirror-Match with Clustering}, two algorithms that we showed to perform better than the state of the art. The novel contributions of the present paper consists of presenting these together with several related algorithms in a general and comprehensive framework.  We further develop the theoretical foundations for comparing the algorithms and use these to formally prove a ranking in terms of performance for the different algorithms, which also enables us to understand why \emph{Mirror-Match} performs better than \emph{Ratio Match} in the first place.  In addition we benchmark all algorithms within the framework extensively on a much larger dataset consisting of over 3000 image pairs.

The paper is organized as follows. In Section~\ref{S:MatchingMethods} we introduce the original \emph{Ratio-Match}, extend on this method to introduce the proposed framework and go on to compare the members of the framework theoretically.  In Section~\ref{S:Experiments} we present evaluations and discuss the results obtained.  Section~\ref{S:Summary} concludes the paper.

@@

To make further discussion easier we introduce the following nomenclature:
\begin{itemize}
\item{Let $f_q$ be any feature point in the \emph{query image} that we 
    are currently trying to match}.
\item{Let $F$ be a set of features. $F_{t}$ for example is the set 
    of features from the \emph{target image}}.
\item{Let $\tau \in [0 \ldots 1]$ be a threshold used to decide whether 
    to keep a match or discard it.} %We keep a match only if it has a \emph{uniqueness ratio} lower than $\tau$}
\item{Let the \emph{proposed match} be the nearest neighbor of the query 
    feature $f_q$ picked from a set of feature points that we will call 
the \emph{proposal set} which does not contain the query feature itself}.
\item{Let the \emph{baseline match} be the nearest neighbor of the query 
    feature $f_q$ picked from a set of feature points that we will call 
the \emph{baseline set} which neither contains the \emph{proposed match} 
nor the query feature}.
\end{itemize}

@@

Given two images we can construct the proposal and baseline sets in six different ways as illustrated in Figure~\ref{fig:overview}. In the case of \emph{Ratio-Match} \cite{lowe2004sift} the \emph{proposal set} consists of all the feature points in the \emph{target image}.  Similarly the \emph{baseline set} consists of all the feature points in the \emph{target image} except the \emph{proposed} match. For \emph{Ratio-Match} this practically means that given a query feature, $f_q$, the \emph{proposed match} and the \emph{baseline match} are the two nearest neighbors in the set of feature points in the \emph{target image}, and the ratio is decided uniquely based on the \emph{target image}. In turn \emph{Mirror-Match} as introduced by the authors in \cite{arnfred2013mirror} works by constructing the \emph{proposal set} and \emph{baseline set} by combining features from both the \emph{query} and \emph{target image}. By similar variations of the \emph{proposal} and \emph{baseline set} we derive four more algorithms as illustrated in Figure~\ref{fig:overview}. We will not discuss \emph{Self-Match-Ext} and \emph{Both-Match} further in this paper since we prove them to be equivalent with \emph{Self-Match} and \emph{Mirror-Match} respectively in \ref{A:self} and \ref{A:mirror}.

@@

For some methods it is possible that the \emph{proposal match} is a feature from the \emph{query image}, in which case we discard the match as an untrue correspondence.  It also happens that we encounter correspondences with $r>1$, in which case the match is also discarded.  Take for example the case of \emph{Self-Match} where we might find that the nearest neighbor of a feature in the \emph{target image} is further from the query feature than the nearest neighbor in the \emph{query image}. In this case the match is discarded.

@@

The main distinguishing factor between the algorithms in the framework is the final ratio between the distance of the two nearest neighbors of a query feature which decides if we keep or discard a match. We illustrate this process in Figure~\ref{fig:matching}.

We can define a calculation of this ratio that is common to all algorithms in the framework, which allows us to compare the algorithms theoretically. To do so we we introduce the \emph{uniqueness ratio} $r$, based on the concept of a nearest neighbor.

Given a feature $f_i$ and a set of features $F$, the nearest neighbor of $f_i$ in $F$ is calculated as follows: 
\begin{equation*}
\argmin_{f_j \in F} d(f_i, f_j)
\end{equation*}
Here, $d(f_i, f_j)$ is the distance between feature descriptors.With SIFT and SURF this is the Euclidean distance of the feature vectors \cite{lowe2004sift,bay2006surf}, whereas BRIEF, BRISK, and FREAK use the Hamming distance \cite{leutenegger2011brisk,calonder2010brief,alahi2012freak}.

@@

\begin{enumerate}
    \item{For a query and \emph{target image} we assume that for any point in 
        the \emph{query image} there is at most one real correspondence in the 
    \emph{target image} and no real correspondences in the \emph{query image} itself 
as assumed by \emph{Ratio-Match}.}
    \item{We assume that distances of feature descriptors are well 
            behaved.  More precisely, given $f_q$, a feature from the 
            \emph{query image}, and $f_{match}$, the real correspondence 
            from the set of features in the \emph{target image} 
            $F_{t}$, we assume $\forall f_i \in F_{proposal}, 
            f_i \neq f_{match}: d(f_q,f_{match}) < d(f_q, f_i)$.}
    %\item{We assume that features within one image resemble each other 
            %more than features in different images. That is, given 
            %$f_a,
    %        f_i \in F_1$ and $f_b, f_j \in F_2$ where $(f_i, f_j)$ and 
    %    $(f_a, f_b)$ are true correspondences, and $F_1$ and $F_2$ are 
    %features collected from different images then we expect that 
            %$d(f_a, f_i)% < d(f_a, f_j)$ and $d(f_b, f_j) < d(f_b, 
            %f_i)$}
     \item{For a set of query features $F_{q}$ with a true 
             correspondence in the \emph{target image}, we assume that 
             the distribution of \emph{uniqueness ratios} using $F_{t}$ as 
             the baseline set is similar to using $F_{q}$, since the 
         two images in the case of true correspondences are bound 
     to share part of the same scene.}
    \end{enumerate}

@@

Next we show that \emph{Mirror-Match} is equal or better than \emph{Ratio-Match-Ext} in terms of precision and recall. Consider as before the nearest neighbor $f_{p}$ of a query feature $f_{q}$. When $f_{p}$ resides in the \emph{query image} both algorithms behave alike and discard the match.  However, consider the uniqueness ratio of \emph{Mirror-Match} for the case where $f_{p}$ resides in the \emph{target image}:
\begin{align*}
    r &= \text{r}(f_{q}, F_{p}, F_{b}) \\
        &= \text{r}(f_{q}, F_{p}, F_{q} \cup F_{t})\\
        &= \max( \text{r}(f_{q}, F_{p}, F_{t}), 
    \text{r}(f_{q}, F_{p}, F_{q}) ).
\end{align*}
For a true correspondence the \emph{uniqueness ratio} using $F_{q}$ as a baseline is distributed similarly to the ratio when using $F_{t}$ according to assumption \#3. In this case the algorithm performs like \emph{Ratio-Match-Ext}.  However, for an untrue correspondence with a baseline match in the \emph{query image} which is closer than the baseline match in the \emph{target image}, \emph{Mirror Match} will return a worse \emph{uniqueness ratio}. This in turns means that \emph{Mirror Match} is equal or better than \emph{Ratio-Match-Ext} in terms of  \emph{Precision} while maintaining equal \emph{Recall}.

Using a similar procedure we can prove that \emph{Self-Match} is equal to \emph{Self-Match-Ext} and \emph{Both-Match} is equal to \emph{Mirror Match}.  The details can be found in \ref{A:self} and \ref{A:mirror}, respectively.

@@

The first assumption states that there is at most one unique match to every feature point, which is often the case for natural images. However, for the recognition of object classes or in images with repetitive content, a point in the \emph{query image} might have several true correspondences in the \emph{target image}, so using a \emph{baseline match} from the \emph{target image} will lead to a much higher \emph{uniqueness ratio}. However, as long as the \emph{query image} does not contain repetitive objects, we can still use \emph{Self-Match} without loss of precision.

@@

For the experiments we use the SIFT descriptor and keypoint detector with default parameters as implemented in the OpenCV library 2.4.6.  To further test how the framework performs across descriptors, we evaluate SURF \cite{bay2006surf}, BRISK \cite{leutenegger2011brisk}, BRIEF \cite{calonder2010brief}, and FREAK \cite{alahi2012freak} in Section~\ref{label:desc}.

@@

\subsection{Evaluation of Methods}
\label{S:3dobjects}

@@

\begin{figure}[htb]
    \centering
    \begin{subfigure}[t]{0.22\columnwidth}
        \centering
        \includegraphics[width=\columnwidth]{images/Frog_A}
        \caption{Query Image}
        \label{fig:frog_a}
    \end{subfigure}%
    \quad %add desired spacing between images, e. g. ~, \quad, \qquad a 
    %blank line to force the subfigure onto a new line)
    \begin{subfigure}[t]{0.22\columnwidth}
        \centering
        \includegraphics[width=\columnwidth]{images/Frog_B}
        \caption{Auxiliary Image}
        \label{fig:frog_b}
    \end{subfigure}%
    \quad %add desired spacing between images, e. g. ~, \quad, \qquad 
    %a blank line to force the subfigure onto a new line)
    \begin{subfigure}[t]{0.22\columnwidth}
        \centering
        \includegraphics[width=\columnwidth]{images/Frog_C}
        \caption{Target Image}
        \label{fig:frog_c}
    \end{subfigure}%
    \quad %add desired spacing between images, e. g. ~, \quad, \qquad 
    %blank line to force the subfigure onto a new line)
    \begin{subfigure}[t]{0.22\columnwidth}
        \centering
        \includegraphics[width=\columnwidth]{images/Frog_C_lines}
        \caption{Target Image with epipolar constraints}
        \label{fig:frog_c_lines}
    \end{subfigure}%
    \caption{Creating epipolar constraints based on three source images \cite{moreels2007evaluation}.  (a) \emph{Query image} marked with the position of the feature we are attempting to match. (b) Auxiliary image, taken at the same rotation as the \emph{query image} but from a higher elevation angle. The line going through the image is the epipolar line of the feature point in the \emph{query image}.  The markers indicate all feature points in the image found near the epipolar line.  In c) we show the \emph{target image} which in this case is rotated 45 degrees from the \emph{query image}.  The line overlaid on the image represents the epipolar line corresponding to the feature point shown in the \emph{query image}.  The markers indicate all feature points in the image found near the epipolar line. In d) we show the \emph{target image} again, this time overlaid with the epipolar lines corresponding to all the features shown in the \emph{auxiliary image}. A true correspondence should be found within a small distance of one of the intersections of the line in c) and the lines in d). In this particular case both feature points shown in c) and d) could possibly be a correct correspondence.
}%
    \label{fig:frog}%
\end{figure}%

@@

We evaluate all matching algorithms from our framework on the 3D Objects dataset by matching images at different angular intervals. For each object we pick the \emph{query} image as the image taken at 10 degrees rotation for calibration stability.  We then match this image with the same object turned an additional $\Delta$ degrees, $\Delta \in \{5, 10, \ldots, 60\}$.  For every angle interval we compare images taken under 3 different lighting conditions as provided by the dataset. We include all objects in the database for which photos at 5 degree angle intervals are available except for the ``Rooster'' and ``Sponge'' objects due to image irregularities.

@@

Figure~\ref{fig:all_objects} shows the performance of the different matching methods in our proposed framework for $12$ increasingly bigger angle differences. The results are shown in a precision / recall plot to make it easy to compare performance in terms of precision at similar levels of recall.  For each plot we show the accumulated results on all 3D objects, weighted by the number of possible true correspondences for the particular object. This ensures that each object contributes equally to the final result, despite some objects resulting in disproportionally more matches than others.

\emph{Ratio-Match} and \emph{Self-Match} show similar results, while \emph{Mirror-Match} and \emph{Ratio-Match-Ext} outperform both of them, showing the advantage of composing the \emph{proposed set} of features from both images. \emph{Mirror-Match} fares slightly but consistently better than \emph{Ratio-Match-Ext}. In general we see the largest performance improvements at lower recall and a  convergence of precision at higher recall. This is expected since a higher recall is a direct consequence of a more lenient threshold; as we let the threshold approach $1$, we lose the benefits of thresholding on the \emph{uniqueness ratio}, and all methods start approximating the results of a simple nearest neighbor match. For all but \emph{Ratio-Match}, the features with better matches within the same image are still weeded out, which explains the tail end of \emph{Ratio-Match} at high recall where the other algorithms no longer have results. As we demonstrated in \cite{arnfred2013mirror}, the removal of within-image matches improves performance on image pairs with partial or no overlap. For no overlap we would expect a matching algorithm to reject matches within regions that have no matching counterparts in the other image.

@@

To summarize, the methods perform as predicted on the 3D Objects dataset with \emph{Mirror-Match} in general performing better than \emph{Ratio-Match-Ext}, which in turn outperforms \emph{Ratio-Match}. The experimental results also show that \emph{Self-Match} overall performs equal to or better than \emph{Ratio-Match}.

\subsection{Evaluation of Descriptors}
\label{label:desc}

@@

The descriptors were evaluated on the 15 images  of the 3D Objects dataset pictured in Figure~\ref{fig:3d_objects}, with a fixed angle offset of 25 degrees.  Otherwise the evaluation is performed in the same manner as the general evaluation of 3D Objects in Section~\ref{S:3dobjects}, and the data presented is accumulated over the 15 objects, each set of results weighted by the possible number of correspondences.

The results in Figure~\ref{fig:descriptors} are again in line with the theoretical performance analysis of Section~\ref{S:Proofs}. They also show a clear difference in the variance of matching method performance between SIFT and the rest of the descriptors. The performance of \emph{Mirror-Match} using SIFT shows a clear increase in precision over \emph{Ratio-Match} at equal recall levels. For other descriptors the improvement is much less pronounced.  For SURF, BRISK, and BRIEF we see a small improvement in precision using \emph{Mirror-Match} over \emph{Ratio-Match} and \emph{Ratio-Match-ext}, whereas FREAK yields similar results no matter what algorithm is used.  For all descriptors \emph{Self-Match} performs on par with \emph{Ratio-Match}.

@@

\subsection{Computational Performance}

@@

From an evaluation using images of rotated 3D objects, we have shown that the theoretical conclusions are confirmed by our experimental evaluation, with \emph{Mirror-Match} often outperforming \emph{Ratio-Match} significantly over 3024 image pairs. Finally we show that the theoretical conclusions hold across a variety of popular feature descriptors (SIFT, SURF, BRISK, BRIEF, and FREAK). These performance gains come for free in terms of both computational complexity as well as actual computing time.

With the framework we introduced three simple methods that are all applicable in situations where \emph{Ratio-Match} is currently used.  \emph{Ratio-Match-Ext} and \emph{Mirror-Match} can readily and easily replace \emph{Ratio-Match} without any performance penalty and with an increase in precision.  However, both are limited to matching image pairs like \emph{Ratio-Match}. \emph{Self-Match} on the other hand, which is on par with \emph{Ratio-Match} across all evaluations, can be applied when matching multiple overlapping images.
